Factors affecting health sector expenditure in OECD countries

### 2025-04-19 products/manuscript/Manuscript.qmd
project #5

This project uses HTML as output format.

@@

-   Prasanga Paudel$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)

@@

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: prasanga.paudel@email.com

@@

_We used Ordinary Least Squares method to analyze the factors affecting per capita health expenditure of a country among the OECD nations during 2010 to 2020. We found that corruption GDP growth, control of corruption, and  unemployment rate have a significant and an important role in determining the amount of money spent on health sector by the country. Foreign Direct Investment had a significant but low magnitude impact. Factors such as income share held by the lowest 20-percent population, trade as a percentage of GDP, access to electricity among rural popuation, and age dependency ration had no statistical significant impact on the health expenditure._

@@

## Description of data and data source

This is a time series data based on 38 member countries from the Organisation for Economic Co-operation and Development (OECD). This organization is a unique forum with governments from 37 democracies collaborate to develop policy standards to promote sustainable economic growth. However, the data obtained from this dataset is crude and needed to be treated before attempting analysis based on it. The details about the data cleaning process can be assessed from this github repository.

@@

*The main objective of this research is to:*

*1. To analyze the factors that affect the expenditure in health sector among OECD countries.*

*2. To analyze the magnitude of the influence of these factors on health expenditure*

@@

## A look into the train dataset

@tbl-summary shows the values under interesting variables for all OECD countries for the year 2010.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of variables"
#| echo: FALSE
table_html <- readRDS(here::here("results", "tables", "health_exp_2010.rds"))
table_html
```

@@

In exploratory data analysis, we will just look at data from Australia as a reference for relation between variables in most of the cases. We are currently thinking of using "Current Health Expenditure as a % of GDP" as our dependent variable. But, we will see it's distribution and other alternatives before making decision.

## Exploratory/Descriptive analysis

*Here, we will now use a combination of tables and figues to have a look into the data.*

@@

### Looking into the trend for our possible dependent variables

@fig-health-expenditure-plot shows how "Health Expenditure as a Percentage of GDP Over Time (2010–2020)" is distributed over the years for top 5 OECD countries.

```{r}
#| label: fig-health-expenditure-plot
#| fig-cap: "Current Health Expenditure as a % of GDP (2010–2020): Top 5 nations"
#| echo: FALSE
knitr::include_graphics(here("results","figures","health_expenditure_plot.png"))
```

But, before moving on to be fixed at a single variable as our dependent variable, we will also look if there are any alternative dependent varibles that could prove to be more useful.

```{r}
#| label: fig-per-capita-health-expenditure-plot
#| fig-cap: "Per Capita Current Health Expenditure (2010–2020): Top 5 nations"
#| echo: FALSE
knitr::include_graphics(here("results","figures","per_capita_health_expenditure_plot.png"))
```

@fig-per-capita-health-expenditure-plot shows the distribution of "Per Capita Health Expenditure of top 5 nations between 2010-2020. This variable shows more variability across the countries, therefore we will go forward with "Per Capita Health Expenditure" as the factor that will represent the spending of different countries in health sectors.

# Basic statistical analysis

## Preparing a model

To help finalize our model, we will first try to look at the relation between our dependent variable "Health Expenditure Per Capita \$USD" with other variables from a specific category. Specifically, we will pick the most useful variables from the large pool of 1400+ variables in our dataset, and we will then categorize these variables based on the aspects they cover. For example, if there is avariable named "Annual GDP growth", we will categorize it as an "Economic" variable. Similarly, if there is a variable named "Unemployment rate" or "Total population", we will categorize them as "Social" variable

We will then run our initial regression with "Health Expenditure Per Capita \$USD" as dependent variable, and variables from each of the category as independent variables in separate regressions

@tbl-regression-results shows the different regresion results based on different categories. The categories are: Economic, Health, Social, Government, Trade. We will pick the most useful variables among the variables that came up significant and create our preliminary model for our research. However, this would not be the end of our variable selection process.

```{r}
#| label: tbl-regression-results
#| tbl-cap: "Regression Results by Category"
#| echo: FALSE
htmltools::tags$iframe(
  src = here("results", "tables", "regression_results.html"),
  width = "100%",
  height = "600px"
)
```

We can observe that health related variables are either missing observations or not statistically significant. But, it should not cause us much trouble because even if we had good observation within health related variables, there would be an equal chance of endogeniety and simultaneity bias with the dependent variable, leading to their exclusion from the model.

## Observing the relationship with dependent variable

@fig-correlation-p1 show the correlation of different variables that had a statistical significant effect in @tbl-regression-results (and other variables of interest). We are just trying to see the pattern present in the dataset for the year 2010, with all the countries treated as cross-sectional units.

@@

We do the same with few other variables.

@@

It can be seen that most of the variables seem to have a high correlation with our variable of interest. Although some variables lack observations, we must recall that this is just for the year 2010, and the whole dataset might have a different distribution pattern.

## Proposed Model for the research {#model}

Based on the informations obtained in @tbl-regression-results, we have selected variable that would best explain our dependent variable. The model is as follow:

@@

We ran the regression based on our primary model described in the [proposed model for the research](#model) section of this article. The results obtained from the regression analysis are presented by @tbl-primary-regression-results.

@@

The model showed a "good" descriptive performance based on it's adjusted R-squared value of 73.4%. The RMSE value for our model (RMSE=1213) is also quite lower compared to the NULL Model which showed a RMSE value of 2364.

We performed the autocorrelation test. The results from the DW test presented by @tbl-DW-test-results confirms the absence of autocorrelation, therefore we can move forward with our model.

```{r}
#| label: tbl-DW-test-results
#| tbl-cap: "Durbin Watson Test Results"
#| echo: FALSE
#| out.width: "25%"
knitr::include_graphics(here("results", "figures", "durbin_watson_test_results.png"))
```

We also performed the Fisher-type Augmented Dickey Fuller test for presence of unit root. The results confirmed that the not all series in our panel data are non-stationary.

```{r}
#| label: tbl-unit-root-results
#| tbl-cap: "Panel Unit Root Test Results"
#| echo: FALSE

@@

@@

We can observe that the variables in general show some common pattern between the two datasets. The difference is acceptable as they are based on entirely different time frames. If we observe the model performance, it is evident that the model performance is weaker than obtined in the train dataset. The RMSE has increased by 383 from 1213 to 1596, however this is still 768 lower than the NULL model. The adjusted R-squared in our test dataset has also dropped by 12.3% to 62%. We also tested Random Forest model as an alternative of the OLS model, and althouth Random Forest showed promising results in our train dataset, the results were worse than OLS model for the test dataset.

@@

Our results shows that annual GDP growth, control of corruption, life expectancy at birth, unemployment rate, and FDI all have a statistically significant effect on the per capita health expenditure of a country. [@Nghiem2017] also found evidence that annual GDP and unemployment rate have a significant effect on the health expenditures, however the variables used in the studies are different. [@Factor2015] has found that country with high level of corruption have a lower spending on health care as a percentage of GDP.

@@

## Conclusions and Policy implications

@@

## Data aquisition

@@

Link to the source: <https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on#>

>>>>

@@

-   Prasanga Paudel$^{1}$ (ORCID: 0000-0000-1234-5678)

@@

@@

_We used Ordinary Least Squares method to analyze the factors affecting per capita health expenditure of a country among the OECD nations during 2010 to 2020. We found that GDP growth, control of corruption, and  unemployment rate among the youth in the labor force have a significant and an important role in determining the amount of money spent on health sector by the country. Foreign Direct Investment had a significant but low magnitude impact. Factors such as income share held by the lowest 20-percent population, trade as a percentage of GDP, access to electricity among rural popuation, and age dependency ration had no statistical significant impact on the health expenditure. We also applied Random Forest model to explain the factors affecting per capita health expenditure of a country. The results of both the OLS model and Random forest was tested using cross-validation and train/test technique, and we found that our OLS model shows a consistent result across all test. Although Random Model was found to have a better explanation capability, the inconsistensies raised a question on it's validity. _

@@

## Data source

This is a time series data based on 38 member countries from the Organisation for Economic Co-operation and Development (OECD). This organization is a unique forum with governments from 37 democracies collaborate to develop policy standards to promote sustainable economic growth. However, the data obtained from this dataset is crude and needed to be treated before attempting analysis based on it. The details about the data cleaning process can be assessed from this [ GitHub](https://github.com/xprasanga/PAUDEL-MADA-project) repository.

@@

The main objective of this research is to:

1. To analyze the factors that affect the expenditure in health sector among OECD countries.

2. To analyze the magnitude of the influence of these factors on health expenditure.

@@

@@

In this section, we will we will explore the data based.

## Descriptive analysis

Here, we will now use a combination of tables and figues to have a look into the data.

@@

### Trends of Development Indicators

In @fig-development-indicators-trends, we will explore differenct variables of interest by observing the average trends across the countries. We can observe that average life expectancy has been increasing consistently through 2010 to 2020. The average physicians per 1000 people shows an upward trend. However, hospital beds show a decreasing trend. The average government effectiveness seems to be dereasing across the countries. The unemployment rate in the OECD countries seems to be consistently decreasing. Finally, if we observe the average Gross Domestic Product(GDP) , we can see that the GDP of OECD countries are somewhat consistent except for the last year 2018.

```{r}
#| label: fig-development-indicators-trends
#| fig-cap: "Trends of Development Indicators (Average across countries)"
#| echo: FALSE
knitr::include_graphics(here("results","figures","development_indicators_trends.png"))
```

### Trends of health expenditure

@fig-combined-health-expenditure-plot shows how "Health Expenditure as a Percentage of GDP Over Time (2010–2020)" and "Current Per Capita Health Expenditure (2010-2020)" are distributed over the years for top 5 OECD countries. We can observe that among the two factors, Health Expenditure as a percentage of GDP is relatively more consistent across the years. Per capita health expenditure across the countries, on the other hand, shows variation across times. It is evident that per capaita health expenditure is more sensitive to external factors and shows immediate response through each years. Our objective is to explain this variation in expenditure.

```{r}
#| label: fig-combined-health-expenditure-plot
#| fig-cap: "Health expenditure across top 5 countries"
#| echo: FALSE
knitr::include_graphics(here("results","figures","combined_health_expenditure_plots.png"))
```

## Basic statistical analysis

### Preparing a model

To help finalize our model, we first tried to look at the relation between our dependent variable "Health Expenditure Per Capita $USD" with other variables from a specific category. Specifically, we picked the most useful variables from the large pool of 1400+ variables in our dataset, and we then categorized these variables based on the aspects they cover. For example, if there is a variable named "Annual GDP growth", we will categorize it as an "Economic" variable. Similarly, if there is a variable named "Unemployment rate" or "Total population", we will categorize them as "Social" variable

We then ran our initial regression with "Health Expenditure Per Capita $USD" as dependent variable, and variables from each of the category as independent variables in separate regressions. The regression results for this preliminary test can be observed under this [ GitHub](https://github.com/xprasanga/PAUDEL-MADA-project) repository under supplementary-materials.

After observing the relation of almost 60 important variables with our variable of interest "Health Expenditure Per Capita $USD", we filtered few variables worth pursuing, and we will discuss them in the sections that will come.

### Observing the relationship with dependent variable

@fig-correlation-p1 shows the correlation of per capita health expenditure with different variables for the year 2010. This fugure shows the pattern present in the dataset for the year 2010, with all the countries treated as cross-sectional units. We are only using the information from one year, year 2010, because visualization across years can be quite complex as we would also have to consider for the trends of each country across years.

@@

@fig-correlation-p1 shows that income share held by lowest 20% of the population is positively correlated with the per capita health expenditure. Similary, estimate of control of corruption and life expectancy at birth also show positive correlation. Annual GDP growth on the other hand shows a low but negative correlation.

We will check few other relation in @fig-correlation-p2.

@@

It can be seen that most of the variables seem to have a moderate and positive correlation with our variable of interest. Although some variables lack observations, we must recall that this is just for the year 2010, and the whole dataset might have a different distribution pattern.

### Proposed Model for the research {#model}

Based on the informations obtained from our preliminary results, we have selected variable that would best explain our dependent variable. The model is as follow:

@@

### Regression Results

We ran the regression based on our primary model described in the section [ 5.2.3](#model) of this article. The results obtained from the regression analysis are presented by @tbl-primary-regression-results.

@@

The model showed a "good" descriptive performance based on it's adjusted R-squared value of 74.3%. The RMSE value for our model (RMSE=1213) is also quite lower compared to the NULL Model which showed a RMSE value of 2364.

#### Autocorrelation Test

We performed the Durbin Watson autocorrelation test for our model. The results from the DW test, presented by @tbl-DW-test-results, confirms the absence of autocorrelation, therefore we can move forward with our model.

```{r}
#| label: tbl-DW-test-results
#| tbl-cap: "Durbin Watson Test Results"
#| echo: FALSE
#| out.width: "40%"
knitr::include_graphics(here("results", "figures", "durbin_watson_test_results.png"))
```

#### ADF unit root test

We also performed the Fisher-type Augmented Dickey Fuller (ADF) test for presence of unit root. The results confirmed that the not all series in our panel data are non-stationary.

```{r}
#| label: tbl-unit-root-results
#| tbl-cap: "ADF Unit Root Test Results"
#| echo: FALSE

@@

Since there is no significant presence of correlation and unit root, we will claim our model to be a valid model. Next, we will test the robustness of our OLS results throuh various machine learning techniques.

#### Cross-Validation of OLS results

In this section, we will observe how our model will perform across different subsets of datasets within our train dataset.As this is time-series dataset, we will consider the fact that the values across different years are correlated with each other. We, therefore, use three different subset of years with five years within each subset as training data within the train dataset. We will use the year after the training dataset as a testing dataset. Therefore, we ended up with three subsets of training data and three testing data.

```{r}
#| label: tbl-OLS-CV-results-full
#| tbl-cap: "OLS cross-validation results "
#| echo: FALSE

htmltools::tags$iframe(
  src = here("results", "tables", "OLS-CV-results-full.html"),
  width = "100%",
  height = "200px",
  scrolling = "auto",
  frameborder = "0"
)
```

It can be observed that the RMSE values across the different combinations of train/test dataset show a somewhat consistent result. The RMSE value for the primary dataset was 1213. The results from CV, as observed in @tbl-ols-cv-results, show even less RMSE in some cases. This is a good thing as our model performs consistently better in cross-validation.

```{r}
#| label: tbl-ols-cv-results
#| tbl-cap: "OLS cross-validation results summary"
#| echo: FALSE

htmltools::tags$div(
  style = "display: flex; justify-content: center;",
  htmltools::tags$iframe(
    src = here("results", "tables", "OLS-CV-results.html"),
    width = "40%",
    height = "110px",
    scrolling = "auto",
    frameborder = "20"
  )
)
```

If we observe the mean RMSE across the three cross-validation subsets, the mean is lower than our primary dataset with a small standard deviation.

### Random Forest model as an Alternative

Next, we will perform our analyses with a different technique other than ordinary least squares method. We will use a Random Forest (RF) model to explain per capita health expenditure of a country. we will use a tree size of 300, and mtry value equal to 3.

#### Random Forest Results

@tbl-rf-importance-results represents the importance factor obtained from the random forest model. It can be observed that the variables that were found to have a significant and high impact on the per capita health expenditure in the OLS model also show a high importance in the random forest model as well, although results are different from OLS as well.

```{r}
#| label: tbl-rf-importance-results
#| tbl-cap: "RF importance plot"
#| echo: FALSE
#| out.width: "100%"
knitr::include_graphics(here("results", "figures", "rf_importance_plot.png"))
```

#### Performance of RF model

The performance of our random forest model is quite exciting. The R-squared value obtained is quite high (0.988. The RMSE value is also 288, which is almost 1000 less than the RMSE value (1213) obtained from our OLS model. This shows that RF can explain our data way more appropriately than an OLS model.

However, we need to test if this result is consistent across various dataset.

```{r}
#| label: tbl-rf-train-perform
#| tbl-cap: "Random Forest TRAIN dataset performance"
#| echo: FALSE

htmltools::tags$iframe(
  src = here("results", "tables", "rf_train_performance.html"),
  width = "100%",
  height = "200px",
  scrolling = "auto",
  frameborder = "0"
)
```

#### Cross-validation of RF model

In this section, we will try to see how our RF performs across various cross-validation datasets. We will perform cv based on the same combination of datasets we used for our OLS model. There are three different subsets of training dataset with five years in each dataset. Also, there will be a one year test dataset for the year immediately coming after the last year of testing dataset.

```{r}
#| label: tbl-rf-cv-results
#| tbl-cap: "Random Forest CV Results"
#| echo: FALSE

htmltools::tags$iframe(
  src = here("results", "tables", "rf_cv_results.html"),
  width = "100%",
  height = "300px",
  scrolling = "auto",
  frameborder = "0"
)
```

The results obtained from the cross-validation as observed in @tbl-rf-cv-results show that the RMSE values have more than doubled in the cross-validation sets.Infact, the highest RMSE value observed in cv is as high as 2.6 times more than the RMSE obtained in the primary train dataset.

```{r}
#| label: tbl-rf-cv-summary
#| tbl-cap: "Random Forest CV Summary"
#| echo: FALSE

htmltools::div(
  style = "text-align: center;",
  htmltools::tags$iframe(
    src = here("results", "tables", "rf_cv_summary.html"),
    width = "100%",
    height = "120px",
    scrolling = "auto",
    frameborder = "0"
  )
)
```

The mean RMSE, 644, despite showing a standard deviation comparable to the OLS model, has a significantly higher value compared to RMSE (288) obtained with the train dataset. This means that our RF model is performing poorly in the case of different subsections of datasets.

### Visual inspection of the two models in train dataset

Finally, in this section, we will try to visually observe the performance of the two models (OLS and Random Forest) in our train dataset that we discussed and worked on till now.

```{r}
#| label: fig-pred-vs-obs-train
#| fig-cap: "Predicted vs Observed values based on TRAIN dataset"
#| echo: FALSE
knitr::include_graphics(here("results","figures","pred_vs_obs_train.png"))
```

It can be observed in @fig-pred-vs-obs-train that the Random Forest clearly out-performs the OLS model in correctly predicting the outcomes. The datapoints for OLS regression are comparitively very hapazardly distributed compared to the datapoints of RF model.

However, we should remember that although the RF model had an outstanding performance in our primary train dataset with RMSE almost 1000 less than RMSE observed in the OLS model, the cross-validation results showed significant fluctuations, due to which robustness of our RF model can be questioned.

### Test dataset

#### Perfrormance of OLS on test dataset

@@

We can observe that the variables in general show some common pattern between the two datasets. The difference is acceptable as they are based on entirely different time frames. If we observe the model performance, it is evident that the model performance is weaker than obtined in the train dataset. The RMSE has increased by 383 from 1213 to 1596, however this is still 768 lower than the NULL model. The adjusted R-squared in our test dataset has also dropped by 12.3% to 62%.

#### Performance of Random Forest on test dataset

We also ran our Random Forest model with the train test dataset. The performance metrics obtained from the model are presented in @tbl-rf-test-perform.

```{r}
#| label: tbl-rf-test-perform
#| tbl-cap: "Random Forest TEST dataset performance"
#| echo: FALSE

htmltools::tags$iframe(
  src = here("results", "tables", "rf_test_performance.html"),
  width = "80%",
  height = "250px",
  scrolling = "auto",
  frameborder = "0"
)
```

We can observe that the RMSE value has again largely increased. The RMSE from test dataset is now as large as 4.6 times more than the RMSE obtained in the primary train dataset. This value is also double than the mean RMSE obtained from cross-validations. Therefore, there is a high suspicion that our our primary RF model sufferes from overfitting, although it performs relatively better than our OLS model.

But, as of now, as our OLS model showed a consistent and robust result across all our datasets and robustness tests, we can conclude that OLS model should be preferred over the Random Forest model despite RF performs better than OLS. Our conclusion is that the RF model can not be trusted because of it's inconsistency and suspicious results.

#### Visual inspection in test dataset

Finally, we will visually observe the performance of both the models in the test dataset. @fig-pred-vs-obs-test shows the distribution of observed values and predicted values using test dataset for both the models.

```{r}
#| label: fig-pred-vs-obs-test
#| fig-cap: "Predicted vs Observed values based on TEST dataset"
#| echo: FALSE
knitr::include_graphics(here("results","figures","pred_vs_obs_test.png"))
```

As expected based on the RMSE values, the RF model does perform better than our OLS model, However, due to inconsistency, we will prefer our OLS model over the RF model.

One thing that should be noted is that the RF model should be optimized and formulated in an efficient manner because it clearly outperformed the OLS model in every comparision across different datasets despite the inconsistency.

@@

Our results shows that annual GDP growth, control of corruption, life expectancy at birth, unemployment rate, and FDI all have a statistically significant effect on the per capita health expenditure of a country. [@Nghiem2017] also found evidence that annual GDP and unemployment rate have a significant effect on the health expenditures, however the variables used in the studies are different. [@Factor2015] has found that country with high level of corruption have a lower spending on health care as a percentage of GDP. [@Giammanco2019] have found that FDI is positively related to factors related to public health. [@Wang2015] argues that the impact of health expenditures on economic growth is somewhat ambiguous.

@@

# Conclusions and Policy implications

@@

# Data aquisition

@@

The dataset can be assessed through this [link.](https://databank.worldbank.org/source/world-development-indicators/Type/TABLE/preview/on#)
